---
title: "Portfolio 7"
author: "Sam Bowyer"
output:
  pdf_document: default
  html_document:
    df_print: paged
urlcolor: blue
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

# Matrices

Matrices are an extremely important data structure in `R` so it will be instructive to examine their usage in some detail through this portfolio.
In particular we will separately discuss dense matrices (which we are already familiar with) and sparse matrices.

## Dense Matrices
As previously seen we can create a matrix by specifying its entries by columns as follows:
```{r}
a = matrix(1:12, 4, 3)
a
```
However, we can also specify the values by row if we set `byrow=T`:
```{r}
b = matrix(1:12, 4, 3, byrow=T)
b
```
The `4` and `3` in here specify the dimensions of the matrix and can be recovered using the `dim` function:
```{r}
dim(a)
```
We can also give the columns and rows names, either at creation of the matrix or afterwards:
```{r}
c = matrix(1:15, 3, 5, dimnames=list(c('X', 'Y', 'Z'), c('1', '2', '3', '4', '5')))
c
rownames(c)
colnames(c)
rownames(c) = c('A','B','C')
c
```
Rows and columns can be obtained as follows: 
```{r}
c[1,]
c[,1]
```
Note that by default this returns a vector for the column, we can use the `drop=F` indexing option to return a matrix instead:
```{r}
c[,1,drop=F]
```

Matrices come with many features in base `R` including transposition (via `t(...)`) matrix multiplication (via `%*%`), inversion (via `solve(...)`) and eigen-decomposition (via `eigen(...)`), for example:
```{r}
d = t(a) %*% b
d

e = matrix(c(-1, 1.5, 1, -1), 2, 2)
e
e_inverse = solve(e)
e_inverse
eigen(e)
```

Finally we also note that `R` has a `Matrix` package that greatly expands the functionality possible with matrices, for instance it allows us to easily get the rank of a matrix with `rankMatrix` (note that we are generating these matrices with the capitalised `Matrix` function and no longer the lowercase `matrix` function):
```{r}
library(Matrix)
f = Matrix(c(2,4,2,4), nrow=2, ncol=2)
g = Matrix(c(2,4,2,5), nrow=2, ncol=2)
c(rankMatrix(f),rankMatrix(g))
```

## Sparse Matrices
Matrix operations are relatively fast in `R` and can be sped up even further by using C/C++ code (with the `Rcpp` package) alongside careful memory management, however, for very large matrices computation time can easily become intractable regardless of implementation. 
It is here useful to distinguish the case in which a matrix is large but sparse (i.e. has mostly entries of 0); ordinarily large matrices would be difficult to work with efficiently but utilising the sparsity present we can improve the computational efficiency by not storing each of the 0 entries separtely.
For this we'll use the `Matrix` package again, but this time with the argument `sparse=TRUE`.
```{r}
h = matrix(c(1, rep(0,999)), 10, 100)
h_sparse = Matrix(c(1, rep(0,999)), nrow=10, ncol=100, sparse=TRUE)
h[1:5,1:5]
h_sparse[1:5,1:5]
c(object.size(h), object.size(h_sparse))
```
We can see here that the sparse matrix requires a lot less storage than the dense matrix, but note that there are storage costs involved in the sparse matrix that would result in higher storage requirements for small, sparse matrices:
```{r}
i = matrix(c(1, rep(0,24)), 5, 5)
i_sparse = Matrix(c(1, rep(0,24)), nrow=5, ncol=5, sparse=TRUE)
i
i_sparse
c(object.size(i), object.size(i_sparse))
```
In this example we can also see that `i_sparse` is of class `ddiMatrix` whereas `h_sparse` was of class `dgCMatrix`.
These refer to the different methods of compression used for storing these matrices--in particular, `dgC` refers to "digits" (rather than `l` for "logicals" for example) stored in a "general" (as opposed to e.g. triangular, symmetric or diagonal) sparse matrix with storage based on the "columns" (instead of, say, `r` for "rows" or `t` for "triplets").
The class `ddiMatrix` comes from the fact that `i_sparse` is a diagonal matrix (which extends from the sparse matrix class).
It makes sense then that we can use a variety of different sparse matrix types, such as `dgRMatrix` for digit matrices stored in compressed sparse row (CSR)---rather than compressed sparse column (CSC)---format.
Furthermore we can often (though not always) convert between these different formats: we can convert from `dgCMatrix` to `dgTMatrix` (and vice versa), but we cannot convert from `dgCMatrix` or `dgTMatrix` to `dgRMatrix`.
Finally, note that we can perform regular matrix operations with sparse matrices and that these will (hopefully) result in a speed up of computation time (though be wary of taking inverses of sparse matrices as the inverses may not themselves be sparse).
```{r}
j_vals = sample(c(0,1), 100000, replace=TRUE, prob=c(0.95, 0.05))
j = matrix(j_vals, 1000, 1000)
j_sparse = Matrix(j_vals, nrow=1000, ncol=1000, sparse=TRUE)

k_vals = sample(c(0,-1,-2), 100000, replace=TRUE, prob=c(0.8, 0.05, 0.15))
k = matrix(k_vals, 1000, 1000)
k_sparse = Matrix(k_vals, nrow=1000, ncol=1000, sparse=TRUE)

j_sparse[1:5,1:5]
k_sparse[1:5,1:5]

system.time(j%*%t(k))
system.time(j_sparse%*%t(k_sparse))
```